On Optimality of Finite Element Integration

### 2015-12-16 TOMS_Paper/assembly.tex
minor fixes

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main steps: initialization (step 1); scheduling of multilinear operands, to reach quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$. $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}|P|,\ |S_P|{)}$. Intuitively, $c_P$ indicates the number of operations induced by $P$ at every innermost loop iteration once sharing elimination will be applied. Take $P$ such that $c_P$ is minimum.
\item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}. If $|P| < |S_P|$, (i) is applied. If $|P| > |S_P|$, (ii) is applied. Otherwise, $|P| = |S_P|$ means that (i) and (ii) have the same impact on the innermost loop. However, the increase in operation count in the outer loops, $i_l$, may still be different. In particular, if (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$, while if (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$. Note that $n_o$ and $n_s$ capture the number of additions required after factorization of $o$ and $s$. We then select (i) or (ii) such that $i_l$ is minimum. 
\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above). The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
\end{enumerate}
\end{Algo}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one. We prevent this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. The operation count induced by a partition $P$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$. This descends directly from two facts: the assumption that the input expression is in normal form and multilinearity. Therefore, what we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality. We do it by contradiction. Let us assume that taking any $P$ such that $c_{P} > c_{P_{min}}$ would lead to a lower operation count. We first observe that a partition (i.e., its multilinear symbols) depends on one of the two multilinear loops; we then indicate with $P^l$ the fact that $P$ depends on $L_l$, $l \in \lbrace j, k\rbrace$. Assuming $P_{min}^l$, we have two possibilities:
\begin{itemize}
\item likewise, $P^{l}$; that is, both $P_{min}$ and $P$ depend on the same linear loop. This is the simplest case, since swapping the order in which the two partitions are handled would not have any side effects -- there is no clash between all possible different re-factorizations. The assumption is then contradicted because the operation count in the innermost loop is the same irrespective of starting with $P_{min}$ or $P$.
\item $P^{l'}$ and $l' \neq l$. In this case, it is possible that operands (symbols) in $P^{l'}$ are factorized and code motion takes place such that $P_{min}$ ``loses'' some of its operands (symbols). Linearity and the assumption $c_{P} > c_{P_{min}}$ tell us that once $P$ is scheduled, we will have $c_{\Lambda''_I} \geq c_{\Lambda'_I} + 1$; that is, there is at least one extra operation (an addition) due to 
\end{itemize}

>>>>

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$. $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}|P|,\ |S_P|{)}$. Intuitively, $c_P$ indicates the number of operations induced by $P$ at every innermost loop iteration once sharing elimination will be applied. Take $P$ such that $c_P$ is minimum.
%; if multiple partitions have same cost, the preference is for the one with the largest occurrences in the input expression.
\item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}. If $|P| < |S_P|$, (i) is applied. If $|P| > |S_P|$, (ii) is applied. Otherwise, $|P| = |S_P|$ means that (i) and (ii) have the same impact on the innermost loop. However, the increase in operation count in the outer loops, $i_l$, may still be different. In particular, if (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$, while if (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$. Note that $n_o$ and $n_s$ capture the number of additions required after factorization of $o$ and $s$. We then select (i) or (ii) such that $i_l$ is minimum. 
\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above). The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
%TODO : this needs to be applied to all lifted sub-expressions as well
\end{enumerate}
\end{Algo}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one. We prevent this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$. What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality. We do it by contradiction. Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$. In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved. We distinguish two cases.
\begin{itemize}
\item $|P| < |S_P|$. This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place. We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
\item $|P| \geq |S_P|$. The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected. Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion. It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$. Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions). Conversely, if we scheduled $P_{min}$ first, the increase would be -- assuming, without loss of generality, that $|S_{P_{min}}| = n - 1$ -- equal to $2(n - 1) - 1$. 
\end{itemize}

%In total, therefore, 
%If none of the $n$ hoisted sub-expressions include operands in $P_{min}$, we have again $c_{\Lambda''_I} = c_{\Lambda'_I}$. If that is not the case,

%We first observe that a partition (i.e., its multilinear symbols) depends on one of the two multilinear loops; we then indicate with $P^l$ the fact that $P$ depends on $L_l$, $l \in \lbrace j, k\rbrace$. Assuming $P_{min}^l$, we have two possibilities:
%\begin{itemize}
%\item likewise, $P^{l}$; that is, both $P_{min}$ and $P$ depend on the same linear loop. This is the simplest case, since swapping the order in which the two partitions are handled would not have any side effects -- there is no clash between all possible different re-factorizations. The assumption is then contradicted because the operation count in the innermost loop is the same irrespective of starting with $P_{min}$ or $P$.
%\item $P^{l'}$ and $l' \neq l$. In this case, it is possible that operands (symbols) in $P^{l'}$ are factorized and code motion takes place such that $P_{min}$ ``loses'' some of its operands (symbols). Linearity and the assumption $c_{P} > c_{P_{min}}$ tell us that once $P$ is scheduled, we will have $c_{\Lambda''_I} \geq c_{\Lambda'_I} + 1$; that is, there is at least one extra operation (an addition) due to 
%\end{itemize}
